Towards Verifying the Bitcoin-S Library

### 2020-07-22 paper-short/main.tex
work on short version

\keywords{Bitcoin, Scala, Bitcoin-S, Stainless} %TODO mandatory; please add comma-separated list of keywords

@@

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

@@

@@

\begin{abstract}
  We try to verify properties of the Bitcoin-S library, a Scala
  implementation of parts of the Bitcoin protocol. We use the
  Stainless verifier which supports programs in a fragment of Scala
  called \emph{Pure Scala}. Since Bitcoin-S is not written in this
  fragment, we extract the relevant code from it and perform a series
  of equivalent transformations until we arrive at code that we
  successfully verify. In that process we find and fix two bugs in
  Bitcoin-S.
\end{abstract}

@@

For software handling cryptocurrency, correctness is clearly crucial.
However, even in very well-tested software such as Bitcoin Core,
serious bugs occur. The most recent example is the bug found in
September 2018 \cite{cve201817144} which essentially allowed to
arbitrarily create new coins. Such software is thus a worthwhile
target for formal verification. In this work, we set out to verify
properties of the bitcoin-s library with the Stainless verifier.

\paragraph{The Bitcoin-S Library.} The bitcoin-s library is an
implementation of parts of the Bitcoin protocol in Scala
\cite{BitcoinS:website,BitcoinS:github}. In particular, it allows to
serialize, deserialize, sign and validate Bitcoin transactions. The
library uses immutable data structures and algebraic data types but is
not specifically written with formal verification in mind. According
to the website, the library is used in production, handling
significant amounts of cryptocurrency each day
\cite{BitcoinS:website}.

\paragraph{The Stainless Verifier.} Stainless is the successor of the
Leon verifier and is developed at EPF Lausanne
\cite{DBLP:conf/ecoop/BlancKKS13,DBLP:conf/pldi/VoirolKK15,DBLP:conf/pldi/BlancK15}. It
is intended to be used by programmers without training in formal
verification. To facilitate that, it accepts specifications written in
the programming language itself (Scala). Also, it focusses on
counterexample finding in addition to proving
correctness. Counterexamples are useful to programmers while
correctness proofs are not -- correctness is obvious or does not hold,
and often both at the same time.

@@

\paragraph{Outline and Properties to Verify.} In the next section we
try to verify the property that a regular (non-coinbase) transaction
can not generate new coins. We call it the \emph{no-inflation
  property}. Trying to verify it, we uncover and fix a bug in the
bitcoin-s library. We then find that there is too much code involved
that lies outside of the supported fragment to currently make this
verification feasible. So we turn to a simpler property to verify. The
simplest possible property we can think of is the fact that adding
zero satoshis to a given amount of satoshis yields the given amount of
satoshis. We call it the \emph{addition-with-zero property} and we try
to verify it in Section~3. Here as well we see that a significant part
of the code lies outside of the supported fragment. We perform a
  series of equivalent transformations on it until we arrive at code that we
  successfully verify. In that process we find and fix a second bug in
  bitcoin-s.

@@

To better understand the validation of a transaction in bitcoin-s, it
is useful to review how transactions are represented and created.

\paragraph{Creating a Transaction.} To create a transaction, we first
need some coins -- an unspent transaction output. We could load an
actual unspent transaction output from the bitcoin network, but we
create one manually in order to see this process. So we first create
an (invalid) transaction with one output in Figure~\ref{fig:prevtx}.

@@

\begin{figure}
\begin{lstlisting}[style=scala]
  val privKey = ECPrivateKey.freshPrivateKey
  val creditingSPK = P2PKHScriptPubKey(pubKey = privKey.publicKey)

  val amount = Satoshis(Int64(10000))

  val utxo = TransactionOutput(currencyUnit = amount, scriptPubKey = creditingSPK)

  val prevTx = BaseTransaction(
    version = Int32.one,
    inputs = List.empty,
    outputs = List(utxo),
    lockTime = UInt32.zero
  )
\end{lstlisting}

  \caption{Creating a transaction output to spend}
  \label{fig:prevtx}
\end{figure}

We first create a keypair, then a lock script with its public key,
then the amount of satoshis, then a transaction output (\texttt{utxo})
for that amount and locked with that script. Finally we create a
transaction with that output and no inputs. Of course, that is not a
valid transaction, because it creates coins out of nothing. In
particular, \texttt{checkTransaction(prevTx)} returns false, simply
because the list of inputs is empty.

\begin{figure}
\begin{lstlisting}[style=scala]
  val outPoint = TransactionOutPoint(prevTx.txId, UInt32.zero)

  val utxoSpendingInfo = BitcoinUTXOSpendingInfo(
    outPoint = outPoint,
    output = utxo,
    signers = List(privKey),
    redeemScriptOpt = None,
    scriptWitnessOpt = None,
    hashType = HashType.sigHashAll
  )

  val utxos = List(utxoSpendingInfo)

  val destinationAmount = Satoshis(Int64(5000))

  val destinationSPK = P2PKHScriptPubKey(pubKey = ECPrivateKey.freshPrivateKey.publicKey)

  val destinations = List(
    TransactionOutput(currencyUnit = destinationAmount, scriptPubKey = destinationSPK)
  )

  val feeRate = SatoshisPerByte(Satoshis.one)

  val networkParams = RegTest // some static values for testing

  val txBuilderF: Future[BitcoinTxBuilder] = BitcoinTxBuilder(
    destinations = destinations, 
    utxos = utxos,               
    feeRate = feeRate,           
    changeSPK = creditingSPK,    // where to send the change
    network = networkParams      
  )

  val txF: Future[Transaction] = txBuilderF.flatMap(_.sign)

  val tx: Transaction = Await.result(txF, 1 second)
\end{lstlisting}
  \caption{Creating a transaction}
  \label{fig:tx}
\end{figure}

Now that we have a transaction output, we create a transaction to
spend it in Figure~\ref{fig:tx}. First, we need a reference to an
output of a previous transaction, here called
\texttt{outPoint}. Second, we add some information on how to spend
that output, in particular, how to sign the transaction. Now we
assemble the list of unspent transaction outputs (\texttt{utxos}), in our case
just one.

We then set the amount of satoshis that we want to spend. The \texttt{Int64}
class aims to emulate a C data type in Scala, and we will look at it
more closely in the next section.

We then create a lock script (\texttt{destinationSPK}) to receive the
coins, create our list of transaction outputs (\texttt{destinations}),
define the fee rate and set some bitcoin network parameters.

Now we create a transaction builder with those data and we tell it to
start signing the transaction in line 34.

Finally, we get the actual signed transaction. We could serialize it
and send it to the Bitcoin network. We can also pass it to the
checkTransaction function, which will return true.

@@

We submitted this fix together with a corresponding unit test to the
bitcoin-s project in a pull request, which has been merged
\cite{BitcoinS:pull435}.

\paragraph{An Attempt at Verification.} Naively trying Stainless on
the entire bitcoin-s codebase results in many errors -- as was to be
expected. We tried to extract only the code relevant to the
no-inflation-property and to verify that. However, even the extracted
code has more than 1500 lines and liberally uses Scala features
outside of the supported fragment. We tried to transform the code into
the supported fragment, but quickly realized that a better approach is
to first verify a simpler property depending on less code and later
come back to the no-inflation property with more experience. So we now
turn to the addition-with-zero property.

@@

It is of course a crucial property we are verifying here: if zero
satoshis were credited to your account, you would not want your
balance to change! It is also the simplest meaningful property to
verify that we can think of. However, the code involved in performing
the addition of two satoshi amounts in bitcoin-s is non-trivial. The
reason for that is a peculiarity of consensus code: agreement with the
majority is more important than correctness, whatever correctness
might mean. The most widely used bitcoin implementation by far is the
reference implementation Bitcoin Core, written in C++. For consensus
code, bitcoin-s has little choice but to be in strict agreement with
the reference implementation. To achieve that, it implements C-like
data types in Scala and then implements functionality using those
C-like data types. For example, the Satoshis class, which represents
an amount of satoshis, is implemented using the class \texttt{Int64} which
aims to represent the C-type \texttt{int64\_t}.

@@

While this is a potentially serious bug, it turns out that
\texttt{checkResult} is only ever called inside a constructor call for
a number type which contains the intended range check, see lines
32-35. The \texttt{checkResult} function thus can, and should, be
removed entirely. The bitcoin-s developers have acknowledged the bug
and we submitted a pull request to fix it \cite{BitcoinS:pull565}.

@@

\paragraph{Transforming the Code.} We now turn to the list of Scala
features used by the extracted code which are not supported by
Stainless and how to transform the code into the supported fragment.
All transformations are \emph{equivalent} in the sense that if the
addition-with-zero property holds for the transformed code, then it
also holds for the code before the transformation.

@@

The original bitcoin-s code we started from, the extracted code, and
the finally verified code are available in our GitHub repository \cite{verification:github}.

@@

We are happy to see some friendly green verifier output. However,
apart from the bugs we found, the main conclusion from this work is
that we had to non-trivially transform even a very small portion of
the code in order to verify it. At the moment, it is unrealistic to
routinely formally verify properties as part of the bitcoin-s
development process. However, Stainless development has already
progressed (e.g.\ type members are supported in recent versions) and
continues to do so (e.g. self-referential type bounds are on the
roadmap). Some missing features that we identified are presumably very
easy to support, like the message parameter in the require
function. Some other features presumably require more substantial
work, like bitwise operations on integer types.

On the other hand, bitcoin-s uses features that might not be supported
even by future Stainless versions, such as calls to Java code. Here
the bitcoin-s code can hopefully be adapted to accommodate formal
verification.

From our results we conclude that formal verification of bitcoin
libraries in general and bitcoin-s in particular is a worthwhile
endeavour. We are looking forward to verifying more substantial parts
of the code in future work.

>>>>

\keywords{Bitcoin, Scala, Bitcoin-S, Stainless}
% mandatory; please add comma-separated list of keywords

@@

% \supplement{}
% optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

@@

% make paragraph headings bold instead of italics
\renewcommand{\paragraph}{\textbf}%

% The default bold(bx) is too wide
% Choose a nicer narrower bold
\DeclareFontSeriesDefault[rm]{bf}{b}

@@

\begin{abstract}
  We try to verify properties of the Bitcoin-S library, a Scala
  implementation of parts of the Bitcoin protocol. We use the
  Stainless verifier which supports programs in a fragment of Scala
  called \emph{Pure Scala}. Since Bitcoin-S is not written in this
  fragment, we extract the relevant code from it and rewrite it until
  we arrive at code that we successfully verify. In that process we
  find and fix two bugs in Bitcoin-S.
\end{abstract}

@@

For software handling cryptocurrency, correctness is clearly crucial.
However, even in very well-tested software such as Bitcoin Core,
serious bugs occur. The most recent example is the bug found in
September 2018 \cite{cve201817144} which essentially allowed to
arbitrarily create new coins. Such software is thus a worthwhile
target for formal verification. In this work, we set out to verify
properties of the Bitcoin-S library with the Stainless verifier. So
this is a case study in applying the Stainless verifier to existing
real-world code.

\paragraph{The Bitcoin-S Library.} The Bitcoin-S library is an
implementation of parts of the Bitcoin protocol in Scala
\cite{BitcoinS:website,BitcoinS:github}. In particular, it allows to
serialize, deserialize, sign and validate Bitcoin transactions. The
library uses immutable data structures and algebraic data types but is
not specifically written with formal verification in mind. According
to the website, the library is used in production, handling
significant amounts of cryptocurrency each day
\cite{BitcoinS:website}.

\paragraph{The Stainless Verifier.} Stainless is the successor of the
Leon verifier and is developed at EPF Lausanne
\cite{DBLP:conf/ecoop/BlancKKS13,DBLP:conf/pldi/VoirolKK15,DBLP:conf/pldi/BlancK15}.
A distinguishing feature of Stainless is that it accepts
specifications written in the programming language itself
(Scala). Also, it focusses on counterexample finding in addition to
proving correctness. Counterexamples are immediately useful to
programmers, which can not be said about correctness proofs.

@@

\paragraph{Outline and Properties to Verify.} In the next section we
try to verify the property that a regular (non-coinbase) transaction
can not generate new coins. We call it the \emph{no-inflation
  property}. Trying to verify it, we uncover and fix a bug in the
Bitcoin-S library. We then find that there is too much code involved
that lies outside of the supported fragment to currently make this
verification feasible. So we turn to a simpler property to verify. The
simplest possible property we can think of is the fact that adding
zero satoshis to a given amount of satoshis yields the given amount of
satoshis. We call it the \emph{addition-with-zero property} and we try
to verify it in Section~3. Here as well we see that a significant part
of the code lies outside of the supported fragment. We rewrite it
until we arrive at code that we successfully verify. In that process
we find and fix a second bug in Bitcoin-S.

@@

@@

@@

We submitted this fix together with a corresponding unit test to the
Bitcoin-S project in a pull request, which has been merged
\cite{BitcoinS:pull435}.

\paragraph{An Attempt at Verification.} Naively trying Stainless on
the entire Bitcoin-S codebase results in many errors -- as was to be
expected. We tried to extract only the code relevant to the
no-inflation-property and to verify that. However, even the extracted
code has more than 1500 lines and liberally uses Scala features
outside of the supported fragment. We tried to transform the code into
the supported fragment, but quickly realized that a better approach is
to first verify a simpler property depending on less code and later
come back to the no-inflation property with more experience. So we now
turn to the addition-with-zero property.

@@

It is of course a crucial property we are verifying here: if zero
satoshis were credited to your account, you would not want your
balance to change! It is also the simplest meaningful property to
verify that we can think of. However, the code involved in performing
the addition of two satoshi amounts in Bitcoin-S is non-trivial. The
reason for that is a peculiarity of consensus code: agreement with the
majority is the only relevant notion of correctness. The most widely
used bitcoin implementation by far is the reference implementation
Bitcoin Core, written in C++. For consensus code, Bitcoin-S has thus
little choice but to be in strict agreement with the reference
implementation. To achieve that, it implements C-like data types in
Scala and then implements functionality using those C-like data
types. For example, the Satoshis class, which represents an amount of
satoshis, is implemented using the class \texttt{Int64} which aims to
represent the C-type \texttt{int64\_t}.

@@

While this is a potentially serious bug, it turns out that
\texttt{checkResult} is only ever called inside a constructor call for
a number type which contains the intended range check, see lines
32-35. The \texttt{checkResult} function thus can, and should, be
removed entirely. The Bitcoin-S developers have acknowledged the bug
and we submitted a pull request to fix it \cite{BitcoinS:pull565}.

@@

\paragraph{Rewriting the Code.} We now turn to the list of Scala
features used by the extracted code which are not supported by
Stainless and how to rewrite the code in the supported fragment.

All transformations are \emph{equivalent} in the sense that if the
addition-with-zero property holds for the transformed code, then it
also holds for the code before the transformation.

@@

The original Bitcoin-S code we started from, the extracted code, and
the finally verified code are available in our GitHub repository \cite{verification:github}.

@@

We are happy to see some friendly green verifier output. However,
apart from the bugs we found, the main conclusion from this work is
that we had to non-trivially transform even a very small portion of
the code in order to verify it. At the moment, it is unrealistic to
routinely formally verify properties as part of the Bitcoin-S
development process. However, Stainless development has already
progressed (e.g.\ type members are supported in recent versions) and
continues to do so (e.g. self-referential type bounds are on the
roadmap). Some missing features that we identified are presumably very
easy to support, like the message parameter in the require
function. Some other features presumably require more substantial
work, like bitwise operations on integer types.

On the other hand, Bitcoin-S uses features that might not be supported
even by future Stainless versions, such as calls to Java code. Here
the Bitcoin-S code can hopefully be adapted to accommodate formal
verification.

From our results we conclude that formal verification of bitcoin
libraries in general and Bitcoin-S in particular is a worthwhile
endeavour. We are looking forward to verifying more substantial parts
of the code in future work.
